Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-12 paper.tex
Quarter-baked discussion

\section{Discussion}
The introduction of VCF as part of the work of the 1000 Genomes
project~\cite{10002015global} has transformed a fragmented

% A computational format, on the other hand, is optimised
% for calculations, for \emph{using} the data, today. A computational format
% is designed with particular data access patterns in mind, ultimately
% to make a broad range of queries and downstream computations as efficient as
% possible.

% Outline of a possible narrative by paragraph:
% \begin{itemize}
% \item When VCF was introduced as part of the work of the 1000 Genomes
% project the software landscape was a real mess. It has entirely
% succeeded in standardising the output of variant callers, and methods
% have interoperated well for the past decade. As a method of interchange
% following classical file-oriented paradigms it works well enough.
% It is also a good archival format: when compressed with a well established
% method like gzip and combined with the well-documented description of
% the text, we can be sure that files will still be readable decades
% from now.

% \item VCF is holding back large-scale genomics, however, and in particular the
% analysis of population scale genetic variation datasets like UKB and GeL.
% Describe the current state of affairs. Variant datasets are chunked up
% into reasonable sized bits (10s of gigabytes) so that they can be processed
% in parallel. Doing anything with them requires interacting with workflow
% managers. In practise, it is not possible to directly write programs to
% work across multiple large-scale datasets: although the chunks of data
% are given in VCF, orchestrating how the chunks are processed and
% the results combined involves substantial work, involving the workflow
% manager, and the details of how chromosomes are split into chunks.
% Cite examples like the phasing pipelines for UKB from Brownings etc.

% \item Distributed computing is the reality, and increasingly datasets
% are being mandatorily stored in public clouds as part of
% Trusted Research Environments. Cloud object stores are the reality
% because it is much cheaper. POSIX file systems are emulated on top of
% this. Benefits of breaking up data into manageable pieces are that
% compute resources can be localised to the data. Compute resources are
% most effectively obtained as short-lived, small VMs that operate
% as part of a large cluster. Large vertically scaled ``pet'' servers
% are much more expensive, and just not feasible for the current
% scale of data. Lots of other architectural
% benefits from the new reality of cloud-based storage.

% \item Classical methods based around Unix pipelines and streaming
% files between processes on a single machine are therefore a
% significant mismatch to the reality of large-scale genetic variation
% datasets. The majority of methods suggested to make working
% with VCF data more efficient are based on these ideas: they all
% create a single large file and regard distribution of work
% in a cluster as extrinsic. Most output VCF as the result of queries,
% which also assumes that the VCF data can be efficiently processed.
% This is a reasonable assumption for thousands of samples, but not
% for larger datasets.

% \item The key issue here is that VCF is not a \emph{computational}
% format. Contrast VCF with plink, which stores data on disk in a
% form that can be computed on directly. Plink is limited in what
% can be descibed though, does not compress, and is not designed
% to distributed.

% \item The large-scale analysis of genetic variation data does not
% need a more efficient file format, which allows files to be downloaded
% somewhat more quickly, and subsets extracted somewhat more efficiently
% as VCF. It needs a standard storage platform that allows data to be
% processed directly without conversion, subsetted efficiently across
% both the variant and sample dimensions, with first-class support
% for all the necessary additional information (GT values etc),
% and that is suited to the realities of today's cloud-based
% storage and processing facilities.

% \item We suggest that the model used by Zarr is a good choice for this.
% We have shown that chunks of the genotype matrix compressed with standard
% methods are competitive with highly specialised methods for
% genotype data. In reality most of the storage is for non-genotype
% data anyway, where everyone is essentially doing the same thing.
% Columnar storage allows getting just (e.g.) the variant positions
% without reading everything. Another benefit is that by storing
% fields separately in different cloud objects, they can have
% different permissions. So, users can be given differential access
% to read parts of the dataset, and this is managed directly
% by the cloud storage. Give example.

@@

% \item Zarr is not perfect. Discuss limitations around ragged arrays,
% etc. Forthcoming improvements in zarr v3. The whole genotype matrix
% paradigm is challenged by graph genomes as well. While this is
% being worked out, there is still plenty of data to analyse as a
% genotype matrix.

@@

Although the Hadoop ecosystem has been hugely successful
in many domains, it has had a limited impact in genomics.
[List some things that have been tried for e.g variant calling
, like ADAM~\citep{nothaft2015rethinking}].
[Briefly mention two approaches using Hadoop clusters
~\citep{boufea2017managing,fan2020variant}, to not very
exiting results].
Distributed analytics for large-scale data is a mature
techology, with many different approaches. Hadoop~\citep{white2012hadoop}
is the original method, and remains the basis of the
dominant ecosystem. HDFS. Spark [CITATION?] lets us
do more complex analytics. Parquet [CITATION?]
Distributed computation has been the reality in the analysis
of genetic variation for many years,
where data is most often provided as per-chromosome VCFs.
This chunking provides a fairly straightforward way to parallelise
computations, using HPC schedulers to split work per-file.
When more fine-grained parallelism is required, users must
provide exact genome coordinates of the genome regions
as part of their submission scripts. Combining results
is a manual operation.
Workflow engines such as
Snakemake~\cite{koster2012snakemake,molder2021sustainable}
and [TODO mention WDL, Cromwell, etc?] make such analyses
much easier, but there is still a great deal of room for
mistakes.
Per-chromosome VCFs are not practical for biobank scale
projects, however, and variant call data tends to be split
into many (still large) files.
For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
hundreds of files per chromosome.
% This is horrible writing, but the point is important.
While these files provide a natural
unit for distributing work, the details of how they
are split differ and essentially there is
no interoperability among large-scale datasets because the VCFs are
so cumbersome that a separate orchestration layer is required to
access the data in parallel.

>>>>

\section{Discussion}
The introduction of VCF as part of the work of the 1000 Genomes
project~\cite{10002015global} has transformed a fragmented 
landscape of mutually incompatible per-program file formats
into an ecosystem of tools that largely
interoperate~\cite{garrison2022spectrum}. In this 
sense VCF has been a huge success, and the underlying data model
is now deeply embedded in bioinformatics practice. As we have 
illustrated here, the shortcomings of the VCF and BCF file
formats (and more generally \emph{any} row-wise encoding of the 
data model) in terms of computational accessibility are severe,
and there can be little doubt that these file formats 
are incurring a large economic cost and 
holding back progress. 
As we have illustrated in the Genomics England
example here, these shortcomings are most acutely felt in 
large-scale data human datasets but any setting in which we 
wish to extract columns of the data from a row-wise 
VCF format will be inefficient 
when datasets are larger than a few hundred megabytes.

As we have demonstrated, the columnar approach, based on the 
popular Zarr standard and widely used compression libraries, 
can support many different types of data access
patterm efficiently, and has excellent computational accessibility.
In this approach there are some significant departures from 
standard practise, which warrant some discussion here.

Perhaps the largest deviation in the approach that we advocate here 
from standard practise is to decouple our concept of a dataset from
a file. Each VCF is broken up into a set of Zarr arrays, each 
corresponding to a large number of chunks stored in a directory
hierarchy (Zarr also supports other forms of chunk storage 
including single files and key-value stores however; see Methods).
While there are some disadvantages to large numbers of small files,
we argue that the benefits greatly outweigh them, particularly 
in the large, centrally controlled dataset setting. 
Firstly, this approach is inherently parallelisable, and chunks across the 
dataset written and read in parallel without needing complex 
synchronisation and prior layout planning. 
Such parallel write access is very difficult 
in the single-file setting, and one of the key weaknesses 
of the HDF5 format~\citep{folk2011overview} [CITATION with 
parallel troubles from HDF5?]
Secondly, mapping chunks in arrays to individually addressed
keys maps well to large-scale cloud object storage. Indeed,
Zarr is specifically designed for such usage, and 
several petabyte scale datasets have been successfully deployed
and used [citations?].
We have focused here on the classical single-server and 
HPC cluster settings, but a major advantage of Zarr is 
its cloud readiness.
% Probably getting into the weeds too much.
% Some large cloud-hosted VCF datasets are currently stored in 
% filesystems that emulate POSIX semantics on top of such object 
% stores, at a substantial cost in performance. Directly hosting 
% these datasets in object stores would have substantial 
% benefits, both in terms of storage costs and performance.
Thirdly, breaking up fields into different directories (or 
cloud buckets) enables fine-grained access control to the 
dataset. For example, all users may be allowed access to
variant-level data, but only a subset permitted to access
call-level individual data. Such access controls can be 
straightforwardly applied using file or cloud-hosting 
methods on the relevant directores/buckets. Such concerns
are entirely decoupled from the file format.

% This is poorly explained
Along with the move away from the single-file-storage paradigm,
the approach we outline here also suggests a shift from 
the classical Unix pipeline-based processing approach 
(but see below for a discussion of how existing pipelines can
be supported). Datasets flowing through pipelines inherently 
implies creating muliple copies of the data, and when 
multiple terabytes of either unrelated or unchanged data 
get copied several times through these pipes large inefficiencies 
are incurred (see Fig~\ref{fig-whole-matrix-compute}). 
We may think of this strategy as ``copy-oriented'' workflows.
A more efficient way to structure such multilayered computations
is in a  ``mask-oriented'' way. For example, to implement some
complex multilayered call filtering strategy we could imagine sequentially
updating a given per-call mask. Masks are cheap (e.g., the 
standard genotype missingness mask in~\ref{tab-genomics-england-data}
uses less than 1\% of the overall storage).

Zarr provides pragmatic solutions to some of the more pressing 
problems facing the analysis of large-scale genetic variation
data, but it is not a solution to all problems. [large number
of alleles, dataset is basically static, ragged arrays,
strings ... any others? Zarr v3 on the way.]

Nonetheless, at least in the single-shared-copy model currently
used by most large-scale variation datasets, Zarr provides
an excellent basis for building a new generation of tools 
that compute \emph{directly} on the data, efficiently
accessing data by variant and by sample.
The vcfzarr specification and scalable vcf2zarr conversion utility
that we have provided here are the starting points of such 
an ecosystem, but much more would be needed to make a viable 
alternative for everyday bioinformatics workflows. However, a few key
pieces of infrastructure would provide a starting point, and would 
be sufficient to greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England.
Firstly, a \texttt{vcfzarrtools} command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) would immediately speed up 
limited region and field specific queries by orders of magnitude,
and would provide compatibility with existing workflows.
Developing such a utility would be a non-trivial but routine
engineering task.

The second key piece of infrastructure required at this large 
scale concerns distributed computation. Even with more efficient
columnar data representations, computations that required examing 
large volumes of per-call data must be distributed to complete 
in a reasonable amount of time. [Dask, Xarray, sgkit?]
Currently users are responsible for orchestrating such calculations
on clusters, and automation would be a significant boon
for reasearcher productivity. [Hail has demonstrated the value of this?]

% We suggest that the future of VCF is to explicitly define the underlying
% data model, and to decouple it from the current row-wise encodings 
% of the VCF text and BCF binary file formats. This will enable data 
% to flow more freely, and provide scope for much-needed innovation.

@@

@@

% Although the Hadoop ecosystem has been hugely successful
% in many domains, it has had a limited impact in genomics.
% [List some things that have been tried for e.g variant calling
% , like ADAM~\citep{nothaft2015rethinking}].
% [Briefly mention two approaches using Hadoop clusters
% ~\citep{boufea2017managing,fan2020variant}, to not very
% exiting results].
% Distributed analytics for large-scale data is a mature
% techology, with many different approaches. Hadoop~\citep{white2012hadoop}
% is the original method, and remains the basis of the
% dominant ecosystem. HDFS. Spark [CITATION?] lets us
% do more complex analytics. Parquet [CITATION?]
% Distributed computation has been the reality in the analysis
% of genetic variation for many years,
% where data is most often provided as per-chromosome VCFs.
% This chunking provides a fairly straightforward way to parallelise
% computations, using HPC schedulers to split work per-file.
% When more fine-grained parallelism is required, users must
% provide exact genome coordinates of the genome regions
% as part of their submission scripts. Combining results
% is a manual operation.
% Workflow engines such as
% Snakemake~\cite{koster2012snakemake,molder2021sustainable}
% and [TODO mention WDL, Cromwell, etc?] make such analyses
% much easier, but there is still a great deal of room for
% mistakes.
% Per-chromosome VCFs are not practical for biobank scale
% projects, however, and variant call data tends to be split
% into many (still large) files.
% For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
% are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
% hundreds of files per chromosome.
% % This is horrible writing, but the point is important.
% While these files provide a natural
% unit for distributing work, the details of how they
% are split differ and essentially there is
% no interoperability among large-scale datasets because the VCFs are
% so cumbersome that a separate orchestration layer is required to
% access the data in parallel.
